---
title: "Housing Data"
author: "Michael Arango"
date: "4/11/2017"
output:
  html_document:
    highlight: textmate
    theme: yeti
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initializing Packages and Importing Data
```{r init, message = FALSE}
library(gdata)
library(tidyverse)
library(stringr)
library(lubridate)
```

```{r df summary, message = FALSE}
df <- read.xls('./data/AmesHousing.xls', stringsAsFactors = FALSE) #na.strings = "EMPTY"
glimpse(df)
```

We will adopt a standard variable naming convention where all variables are lowercase and words are concatenated with periods.

```{r fix col names}
names(df) <- tolower(names(df))
# drop the order vairiable
df <- df %>% 
  select(everything(), -order)
names(df)
```

We decided to drop the order variable as it is just a duplicate row index and corresponds to the observation number. 

## Description of Dataset
The dataset we chose for our project contains information from the Ames Assessor’s Office on residential property sales that occurred in Ames, Iowa from 2006 to 2010.[^1] Thus, the unit of observation in the dataset is a single property sale in Ames, Iowa in a given year. The dataset contains `r dim(df)[1]` observations and `r dim(df)[2]` variables. 

Of the `r dim(df)[2]` variables, 23 are nominal, 23 are ordinal, 14 are discrete, and 20 are continuous. The variables included are basic characteristics that anyone wanting to buy a house would be interested in. For the most part, the different variables may be split up into specific groups. In general, the 20 continuous variables relate measurements of area dimensions for each observation. These include, among others, the sizes of lots, rooms, porches, and garages. The 14 discrete variables mostly have to do with the number of bedrooms, bathrooms, kitchens, etc. that a given property has. There are several geographic categorical variables that start profiling properties at the individual Parcel ID level and end at the neighborhood level. The rest of the nominal variables identify characteristics of the property and dwelling type/structure. Most of the ordinal variables are rankings of the quality/condition of rooms and lot characteristics. For more information on the variables in the dataset, consult the `DataDescription.txt` file in the data directory. 

[^1]: *Journal of Statistics Education*, Volume 19, Number 3(2011), accessed April 11, 2017 at www.amstat.org/publications/jse/v19n3/decock.pdf

# Cleaning 
```{r missing by col}
# Are there any missing values in the data
any(is.na(df))
# How many are there
sum(is.na(df))
# return index of columns that have missing values 
na.cols = which(colSums(is.na(df)) > 0)
# Break down missing values by variable
sort(colSums(sapply(df[na.cols], is.na)), decreasing = TRUE)
```

One thing to be aware of is that some variables such as `Alley` have `NA` encoded as a level to specify `No Alley Access`. Thus, these cases are not "missing" values. 

```{r plot helper}
# This is a helper function to plot the categorical variables in the dataset
plotCategorical = function(cols, dataframe) {
  for (col in cols) {
    # Remove NA's & sort categories by tally
    order.cols = names(sort(table(df[,col]), decreasing = TRUE))
    # qplot is ggplot's equivalent of base R's high-level plotting function `plot`
    num.plot = qplot(dataframe[,col]) +
      # change bar color 
      geom_bar(fill = 'cornflowerblue') +
      # add the value labels to each bar
      geom_text(aes(label = ..count..), stat='count', vjust=-0.5) +
      # minimal theme
      theme_minimal() +
      # set scales for each plot to go from 0 to max of categorical feature
      scale_y_continuous(limits = c(0,max(table(dataframe[,col]))*1.1)) +
      scale_x_discrete(limits = order.cols) +
      xlab(col) +
      # rotate x-axis label text 30 degrees and set font size to 12
      theme(axis.text.x = element_text(angle = 30, size=12))
     # Show plot and suppress warning messages from plot function
     suppressWarnings(print(num.plot))
  }
}
```

## Parcel ID Number

We already saw that the `pid` variable does not have any missing values, but it might be interesting to see if any parcel appears twice. We would be able to see appreciation/depreciation over time if the same house was bought and sold within a 5 year window. 

```{r pid}
summarise(df,
          n_obs = n(),
          n_pid = n_distinct(pid))
```

Unfortunately, every parcel ID only appears once in the dataset. 

## Dwelling Type

The `ms.subclass` variable specifies the type of dwelling involved in the sale. The variables as coded as an integer and has a key that can be found on the documentation. There are `r n_distinct(df$ms.subclass)` types of dwellings included. Numbers in the 20 t0 85 range represent different traditional style homes and others represent duplexes or other split homes. 

```{r ms.subclass}
table(df$ms.subclass)
```

The table shows that most residential sales during this time were one or two-story houses built after 1946. 

## Pool Quality 
Pool quality is an ordinal description of the condition of the pool. It seems there are so many NA's because the houses are in Ames, Iowa and you cannot rate pool quality if a house does not have a pool. Since we have the actual dimensions of the pools for all houses, we can double-check to make sure if this assumption is correct. 

```{r plot pool.qc}
plotCategorical('pool.qc', df)
```

The following code will return all rows (if there are any) where the pool area is greater than 0 and the pool quality is missing. 

```{r missing pool_qc}
df[(df$pool.area > 0) & is.na(df$pool.qc), c('pool.qc','pool.area')]
```

It appears that no pool quality information is missing for houses that have pools. This checks out as there are 2930 rows, 2917 missing values in the pool quality variable, and 13 recorded pool quality records in the bar chart. We changed the missing values for pool quality from `NA` to `No Pool` to give it an appropriate level. 

```{r}
index <- which(is.na(df$pool.qc))
df[index, 'pool.qc'] <- 'No Pool'
```











```{r}
#ggplot(data=df, aes(x=gr.liv.area, y=saleprice)) + geom_point(alpha=.25, color="#00b9be")
```
