---
title: "Housing Data"
author: "Michael Arango"
date: "4/11/2017"
output:
  html_document:
    highlight: textmate
    theme: yeti
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initializing Packages and Importing Data
```{r init, message = FALSE}
library(gdata)
library(tidyverse)
library(stringr)
library(lubridate)

```

```{r dfsummary, message = FALSE}
df <- read.xls('./data/AmesHousing.xls', stringsAsFactors = TRUE)
glimpse(df)
```

## Description of Dataset
The dataset we chose to use for our project contains information on residential property sales that occurred in Ames, Iowa from 2006 to 2010.[^1] Thus, the unit of observation in the dataset is a single property sale in Ames, Iowa in a given year. The dataset contains `r dim(df)[1]` observations and `r dim(df)[2]` variables. 

Of the `r dim(df)[2]` variables, 23 are nominal, 23 are ordinal, 14 are discrete, and 20 are continuous. The variables included are basic characteristics that anyone wanting to buy a house would be interested in. For the most part, the different variables may be split up into specific groups. In general, the 20 continuous variables relate measurements of area dimensions for each observation. These include, among others, the sizes of lots, rooms, porches, and garages. The 14 discrete variables mostly have to do with the number of bedrooms, bathrooms, kitchens, etc. that a given property has. There are several geographic categorical variables that start profiling properties at the individual Parcel ID level and end at the neighborhood level. The rest of the nominal variables identify characteristics of the property and dwelling type/structure. Most of the ordinal variables are rankings of the quality/condition of rooms and lot characteristics. For more information on the variables in the dataset, consult the `DataDescription.txt` file in the data directory. 

[^1]: *Journal of Statistics Education*, Volume 19, Number 3(2011), accessed April 11, 2017 at www.amstat.org/publications/jse/v19n3/decock.pdf





