---
title: "Data_Analysis_Yesh"
author: "Yeshwant Chillakuru"
date: "4/27/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadData, message=FALSE}
library(tidyverse)
library(pls)
library(stargazer)

df <- read.csv("../train_housing.csv", header=TRUE)
df_test <- read.csv("../test_housing.csv", header=TRUE)
```

# Remove all factor columns from data
```{r factordrop}
df_num <- df

for(i in colnames(df)){
  if(is.factor( df[[i]] )){
    df_num[[i]] <- NULL
  }
}

str(df_num)
```


# PCR Analysis

## Model Selection
```{r pcr1}

pcr.saleprice <- pcr(saleprice ~ ., data=df_num, scale=TRUE, validation="CV")

validationplot(pcr.saleprice)
summary(pcr.saleprice)
```

We can see that 1 component explains about 80% of the variance in `saleprice`. This rises only marginally afterwards. The next big jump is from 6 to 7 components, and again from 23 to 24. 3 Models will be assessed with 1, 7, and 24 components.

# Model Validation
```{r modelvaldiation}

pcr.predict.1 <- predict(pcr.saleprice, df_test, ncomp=1)
df_predict.1 <- data.frame(pcr.predict.1, df_test$saleprice)
saleprice_MSE.1 <- mean((df_predict.1[,1]-df_predict.1[,2])^2)
saleprice_RMSE.1 <- sqrt(saleprice_MSE.1)

pcr.predict.2 <- predict(pcr.saleprice, df_test, ncomp=7)
df_predict.2 <- data.frame(pcr.predict.2, df_test$saleprice)
saleprice_MSE.2 <- mean((df_predict.2[,1]-df_predict.2[,2])^2)
saleprice_RMSE.2 <- sqrt(saleprice_MSE.2)

pcr.predict.3 <- predict(pcr.saleprice, df_test, ncomp=24)
df_predict.3 <- data.frame(pcr.predict.3, df_test$saleprice)
saleprice_MSE.3 <- mean((df_predict.3[,1]-df_predict.3[,2])^2)
saleprice_RMSE.3 <- sqrt(saleprice_MSE.3)

stargazer(df_predict.1, df_predict.2, df_predict.3, type="text")
```

Comparing the 3 Models, we see the following RMSE Values:

  * Model 1 (1 Component): `r paste(round(saleprice_RMSE.1))`
  * Model 2 (7 Component): `r paste(round(saleprice_RMSE.2))`
  * Model 3 (24 Component): `r paste(round(saleprice_RMSE.3))`

Model 2 appears to be the best, as Model 3 risks overfitting for not a not much better fit.















